---
title: "Introduction to Cartesian Behaviour"
author: "Faco (de Souza, Falco Luís Gonçalves)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro_Cartesian_Behavior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r setup, include = FALSE}

# Definir mirror CRAN globalmente para garantir que install.packages() funcione ao knit
options(repos = c(CRAN = "https://cloud.r-project.org"))
# Instalar pacotes faltantes, se necessários
pkgs <- c(
  "tidyverse", "knitr", "rmarkdown", "htmlTable",
  "kableExtra", "CartesianBehaviour", "trajr"
)
to_install <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly=TRUE)]
if (length(to_install)) {
  install.packages(to_install, repos = getOption("repos"))
}
# Configurações de chunk
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 6.5, fig.align = "center",
  cache = TRUE
)

library(tidyverse)
library(htmlTable)
library(kableExtra)

```

## Briefing introduction
<br>
This package aims to simplify the way to deal with Cartesian data and turn it into movement and maybe behavioral evidence. Note that this package don't track nor generate the trajectories, for this we recommend [TrackR](https://swarm-lab.github.io/trackR/)[,](https://www.youtube.com/watch?v=JVyPSIiqgwk) [trajr](https://github.com/JimMcL/trajr) and [DeepLabCut](https://github.com/DeepLabCut/DeepLabCut). In this first version you must create zones of interest even if the only zone of interest is the whole tracking area. Also the outputs can be computed as over time data if any interest.
<br>

This vignette demonstrates how to use CartesianBehaviour to convert Cartesian‐coordinate trajectories into zone‐based and time‐series behavioral metrics. You will learn how to:

Generate or import a trajectory.

Define zones of interest (even if it’s just the full arena).

Compute per-zone and over-time summaries for single and multiple tracks.


## 1. Single‐track analysis
### 1.1 Generate a synthetic trajectory

``` {r}
Traj <- trajr::TrajGenerate(
  n = 10000, random = TRUE,
  stepLength = .009, angularErrorSd = .5, fps = 30
)
head(Traj)
```
### 1.2 Define zones of interest
``` {r}

Z1 <- data.frame(x = c(-12, 0, 0, -12), y = c(10, 10, 0, 0))
Z2 <- data.frame(x = c(0, 12, 12, 0),   y = c(10, 10, 0, 0))
Z3 <- data.frame(x = c(0, 12, 12, 0),   y = c(0,  0, -10, -10))
Z4 <- data.frame(x = c(-12, 0, 0, -12), y = c(0,  0, -10, -10))
zones_list <- list(list(Z1, Z2, Z3, Z4))

```
### 1.3 Compute zone‐based metrics
``` {r}

TRAJ2D <- CartesianBehaviour::Allbasics(
  Traj, Xaxi = "x", Yaxi = "y", frames = "time",
  Zones = zones_list, threshold = 3, Dist.reg = "point",
  reg = data.frame(x = 0, y = 0)
)
Distance.Zones <- data.frame(
  Zone = t(CartesianBehaviour::dist_zone(TRAJ2D))
)

```

``` {r, echo = F}
htmlTable(
  round(Distance.Zones, 2),
  pos.caption = "bottom",
  caption = "Total distance traveled per zone (fictional units)."
)
```
Here we plot the data from this single track using [ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html)[.](https://www.youtube.com/watch?v=i5uSbp0YDhc)
``` {r}
#| fig.alt = "Plot of the total distance traveled an te distance traveled inside the zones of interest" 
ggplot(Distance.Zones, aes(x = rownames(Distance.Zones), y = Zone)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black") +
  labs(x = "Zone", y = "Distance", title = "Distance by Zone") +
  ylim(0, max(Distance.Zones$Zone) * 1.1) +
  theme_minimal() +
  theme(
    plot.title   = element_text(hjust = .5, size = 14, face = "bold"),
    axis.title   = element_text(size = 12, face = "bold")
  )

```

## 2. Multi‐track (group) analysis

For any of the measures in this package, if you are dealing with multiple samples, it is possible to list them:

```
List_of_tracks <- list(track1,track2,track3, .... trackN)
```
And call **g_** + **function**. It will operate the samples the same way in a single track data frame.<br>


### 2.1 Generate multiple trajectories
``` {r}

set.seed(42)
TRAJ_list <- replicate(10,
  trajr::TrajGenerate(10000, random = TRUE, stepLength = .01, angularErrorSd = .8, fps = 30),
  simplify = FALSE
)

```
### 2.2 Compute group measures
``` {r, results='hide'}

TRAJS2D <- CartesianBehaviour::g_Allbasics(
  TRAJ_list, Xaxi = "x", Yaxi = "y", frames = "time",
  Zones = zones_list, threshold = 3, Dist.reg = "point",
  reg = data.frame(x = 0, y = 0)
)
Group.Distance.Point <- CartesianBehaviour::g_over_time_ana(
  TRAJS2D, time.ana = 2, type = "point", time.length = 10
)

PTtm <- Group.Distance.Point[, -1]
summary_df <- tibble(
  time = Group.Distance.Point$time,
  mean = rowMeans(PTtm),
  sd   = apply(PTtm, 1, sd)
)
```

``` {r, echo = F}

knitr::kable(Group.Distance.Point, format = "html", align = 'c') %>%
  kable_styling("striped", full_width = FALSE)

```

``` {r echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
#| fig.alt = "Plot of the mean distance to a point over time for muutiple samples with SD"
ggplot(data=summary_df, aes(x=time, y=mean)) +
  geom_smooth(method = NULL)+
  geom_point(size=3) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.5) +
  labs(
    title = "Mean distance over time",
    subtitle = "",
    x = "Time",
    y = "Distance (fictional units)"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = "red",
                              family = "sans", angle = 0, vjust = 0, lineheight = 1),
    axis.title.x = element_text(hjust = 0.5, size = 12, face = "bold", colour = "black",
                                family = "sans", angle = 0, vjust = 0, lineheight = 1),
    axis.text.x = element_text(hjust = 0.5, size = 12, face = "plain", colour = "black",
                               family = "sans", angle = 0, vjust = 0, lineheight = 1),
    axis.text.y = element_text(hjust = 0.5, size = 12, face = "plain", colour = "black",
                               family = "sans", angle = 90, vjust = 0, lineheight = 1),
    axis.title.y = element_text(hjust = 0.5, size = 12, face = "bold", colour = "black",
                                family = "sans", angle = 90, vjust = 0, lineheight = 1),
    plot.subtitle = element_text(size = 10, face = "bold", hjust=0.5),
    plot.caption = element_text(face = "italic"))

```
